---
title: "Figures for PRIMM PhIP-Seq paper"
subtitle: "Plots created with phiper package"
author: "Magdalena Hübner"
license: "CC-BY 4.0"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    theme: united
    highlight: tango
    toc: true              # Enable table of contents
    toc_float: true        # Make the table of contents float on the side
    number_sections: true  # Optional: Number your sections
    toc_depth: 4  
---

<!-- R markdown settings -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE,
  fig.width = 5,
  fig.height = 4
)
```

<!-- Project setup -->

```{r}
# Core tidyverse
library(purrr)
library(stringr)
library(tidyr)
library(tibble)
library(dplyr)

# I/O
library(here)
library(openxlsx)

# Plotting
library(GGally)
library(ggplot2)
library(ggpubr)
library(gt)
library(patchwork)

# PhIP-Seq
library(phiper)
```

```{r}
# Seed (for jitter etc.)
set.seed(612)

# Suppress phiper logging in report
options(phiper.log.verbose = FALSE)
```

```{r}
# Colours scheme for plotting
group_palette <- c(
  # Responders
  "R" = phip_palette[1],
  "R_T0" = phip_palette[1],
  "R_T1" = phip_palette[1],
  # Non-responders
  "NR" = phip_palette[9],
  "NR_T0" = phip_palette[9],
  "NR_T1" = phip_palette[9]
)
```

<!-- Loading functions and pre-computed data -->

```{r}
source(here("R", "helpers.R"))
```

```{r}
# Path to original data from Israel
zip_path <- "/lisc/data/work/ccr/SHARED_RESOURCES/external_data/PRIMM/FinalB_2.zip"
```

```{r}
# PhIP-Seq data
ps <- phip_convert(here("data/processed/", "primm_full.parquet")) %>%
  mutate(group_char = paste(response, timepoint, sep = "_"))

# Sample metadata
meta <- read.csv(here("data/processed/", "metadata.csv")) %>%
  mutate(group_char = paste(response, timepoint, sep = "_"))

# Peptide library
peplib <- readRDS(here("results", "peptide_library.rds"))

# Alpha diversity
alpha_div <- read.xlsx(here("results/R_T0_vs_NR_T0", "alpha_diversity", "table.xlsx"))

# Beta diversity
pcoa_res <- list()
perm_res <- list()
disp_res <- list()
for(cmp in c("R_T0_vs_NR_T0", "R_T1_vs_NR_T1")) {
  pcoa_res[[cmp]] <- readRDS(here("results", cmp, "beta_diversity", "pcoa_results.rds"))
  perm_res[[cmp]] <- readRDS(here("results", cmp, "beta_diversity", "permanova_results.rds"))
  disp_res[[cmp]] <- readRDS(here("results", cmp, "beta_diversity", "dispersion_results.rds"))
}

# POP framework
pep_tbl <- list()
pval_mw_df <- list()
for(cmp in c(
  "R_T0_vs_NR_T0", 
  "R_T1_vs_NR_T1",
  "R_T2_vs_NR_T2", 
  "R_T3_vs_NR_T3",
  "R_T0_vs_R_T1", 
  "NR_T0_vs_NR_T1"
)) {
  pep_tbl[[cmp]] <- readRDS(here("results", cmp, "POP_framework", "single_peptide.rds"))
  pval_mw_df[[cmp]] <- read.csv(here("results", cmp, "POP_framework", "pval_mw_df.csv"))
}
pop_df <- read.csv(
  here("results/R_T0_vs_NR_T0/POP_framework/taxa_ranks.csv"),
  row.names = 1
)

# DELTA framework
delta_df <- read.csv(
  here("results/R_T0_vs_NR_T0/DELTA_framework/delta_table.csv"),
  row.names = 1
)
```

```{r}
# Survival data
meta_surv <- read.csv(here("data/raw/mdat_baseline_survival.csv"))
meta_surv_new <- read.csv(here("data/raw/meta_os.csv"))
```

```{r}
# POP framework (other contrasts)
# Vector of group variables and timepoints
group_vars <- c("ORR", "toxicity", "colitis", "combiIO", "antibiotics", "ppi")
timepoints <- c("T0", "T1", "T2", "T3")
levels2 <- c("yes","no")

# All group labels that to consider
labs <- expand_grid(
  group_type = group_vars,
  group_value = levels2,
  timepoint = timepoints
) %>%
  mutate(
    lab = str_c(group_type, group_value, timepoint, sep = "_"),
    group_lvl = str_c(group_type, group_value, sep = "_")
  )

# 1) within-timepoint: yes vs. no
cmp_within <- labs %>%
  select(group_type, group_value, timepoint, lab) %>%
  pivot_wider(names_from = group_value, values_from = lab) %>%
  transmute(cmp = map2(yes, no, c))

comp_xtra <- cmp_within$cmp
pep_tbl_xtra <- list()
pval_mw_df_xtra <- list()
for(cmp_pair in comp_xtra) {
  cmp <- paste0(cmp_pair[1], "_vs_", cmp_pair[2])
  pep_tbl_xtra[[cmp]] <- readRDS(here("results", cmp, "POP_framework", "single_peptide.rds"))
  pval_mw_df_xtra[[cmp]] <- read.csv(here("results", cmp, "POP_framework", "pval_mw_df.csv"))
}
```


# Metadata

I noticed that there is one sample (**LV1016273557**) which has two metadata
entries belonging to two completely different patients. 

```{r}
meta_all <- read.csv(here("data/processed/", "metadata_all.csv")) %>%
  mutate(group_char = paste(response, timepoint, sep = "_")) %>%
  select(LV_code, patientid, timepoint, centre, age, Sample.ID, response)

meta_all %>% filter(LV_code == "LV1016273557")
meta_all %>% filter(patientid == "MVCC011")
meta_all %>% filter(patientid == "OL_3049")
```

My suspicion is that the sample belongs to OL_3049 (T0). Because otherwise if
it belonged to MVCC011 (T2) this sample should correlate with MVCC011 (T0 & T1)
which is doesn't:

```{r}
checks <- c("T0_MVCC011", "T1_MVCC011", "T0_OL_3049")
hmout_corr <- ps %>%
  mutate(label = paste(timepoint, patientid, sep = "_")) %>%
  filter(label %in% checks) %>%
  select(peptide_id, fold_change, label) %>%
  collect() %>%
  pivot_wider(names_from = "label", values_from = "fold_change") %>% 
  select(all_of(checks))

ggpairs(
  hmout_corr,
  upper = list(continuous = wrap("cor", size = 4, use = "pairwise.complete.obs")),
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.5)),
  diag  = list(continuous = wrap("blankDiag"))
) +
  theme_bw()
```

Initially I removed this sample from the analysis.
This is how the plots for Johannes grant were generated.

```{r}
# ps <- ps %>% filter(sample_id != "LV1016273557")
```


# Figure 1 {.tabset}

## Alpha Diversity {.tabset}

- Richness: The absolute number of antibody-bound peptides/taxa rank per 
individual between groups.

- Shannon: A given taxonomic rank diversity of Ig repertoires of antibody-bound 
  peptides per individual between groups.
  
**This was done with phiper**

We could consider using a richness cutoff (e.g. 200)?

```{r, fig.width=7, fig.height=3.5}
# Alpha diversity plot (Richness + Shannon)
metrics <- c("richness", "shannon_diversity")
cmp_alpha <- c("R_T0", "NR_T0")

p_alpha <- lapply(metrics, function(m) {
  # Calculate and format p-values first
  pv <- compare_means(
    formula = reformulate("group_char", response = m),
    data = alpha_div,
    method = "wilcox.test",
    comparisons = list(cmp_alpha)
  ) %>%
    mutate(
      p = format_pval(p),
      y.position = max(alpha_div[[m]], na.rm = TRUE) * 1.1
    )

  p_alpha <- phiper::plot_alpha_diversity(
    alpha_div,
    metric = m,
    group_col = "group_char",
    x_order = cmp_alpha
  ) +
    stat_pvalue_manual(
      pv,
      label = "p",
      y.position = "y.position",
      vjust = -0.3, # increase space between bracket and text
      tip.length = 0.02,
      label.size = 4.23,
      family = "Montserrat"
    ) +
    scale_fill_manual(values = group_palette) +
    coord_cartesian(clip = "off") +
    theme(
      text = element_text(size = 12),
      axis.title.x = element_blank(),
      axis.title.y = element_text(face = "plain"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = margin(5.5, 5.5, 5.5, 5.5, unit = "pt")
    )

  if (m == "richness") {
    p_alpha <- p_alpha +
      labs(y = "# of significantly bound peptides") +
      ggrepel::geom_text_repel(
        data = alpha_div %>% filter(richness < 200),
        mapping = aes(label = sample_id),
        size = 3,
        show.legend = FALSE
      )
  }
  p_alpha
})

wrap_plots(p_alpha)
```

## Beta Diversity {.tabset}

- For each peptide Bray–Curtis dissimilarity is computed on the binary abundance 
taxonomic matrix and carried out a Principal Coordinates Analysis (PCoA/MDS). 

- On the scatterplot, each point is a sample and the 95% confidence ellipses 
show the group clustering.

- Tested for overall differences in community composition between groups with 
PERMANOVA, reporting the p-value in the plot. To check that any PERMANOVA result 
wasn’t driven by unequal dispersion, a permutation test on the multivariate 
dispersions was run.

- Together, these panels tell you whether (a) the two groups occupy distinct 
regions of Bray–Curtis space (significant PERMANOVA) and (b) whether one group 
is more dispersed than the other (significant dispersion test).

**This was done with phiper**

I can see 2 clusters at TP1, is this an experimental batch effect?
--> Probably CMV

```{r, fig.width=7, fig.height=3.5}
# PCoA plots
cmp_pcoa <- list(c("R_T0", "NR_T0"), c("R_T1", "NR_T1"))

p_pcoa <- lapply(cmp_pcoa, function(cmp) {
  # Add group information to PCoA sample coordinates
  cmp_lab <- paste(cmp[[1]], "vs", cmp[[2]], sep = "_")
  pcoa_res[[cmp_lab]]$sample_coords <- pcoa_res[[cmp_lab]]$sample_coords %>%
    left_join(
      meta %>% select(LV_code, group_char),
      by = c("sample_id" = "LV_code")
    ) %>%
    mutate(group_char = factor(group_char, levels = cmp)) %>%
    left_join(
      alpha_div %>% select(sample_id, richness),
      by = "sample_id"
    )
  
  lab_perm <- paste0("PERMANOVA p = ", format_pval(perm_res[[cmp_lab]]$p_adjust))
  lab_disp <- paste0( "Dispersion   p = ", format_pval(disp_res[[cmp_lab]]$tests$p_adjust))

  # PCoA plot with group centroids and ellipses
  p_pcoa <- phiper:::plot_pcoa(
    pcoa_res[[cmp_lab]],
    axes = c(1, 2),
    group_col = "group_char",
    show_ellipses = FALSE,
    show_centroids = TRUE,
    point_size = 2
  ) +
    stat_ellipse(
      aes(group = group_char),
      colour = "grey70",
      linetype = "dashed"
    ) +
    scale_colour_manual(values = group_palette) +
    theme(
      text = element_text(size = 12),
      legend.title = element_blank(),
      axis.title = element_text(face = "plain"),
      plot.margin = margin(5.5, 5.5, 5.5, 5.5, unit = "pt"),
      legend.position = "none"
    ) +
    annotate(
      "text",
      x = Inf,
      y = -Inf,
      label = paste(lab_perm, lab_disp, sep = "\n"),
      hjust = 1,  # nudge left a bit
      vjust = -0.2,  # nudge up a bit
      size = 4.23
    )
})

wrap_plots(p_pcoa)
```

## Heatmap {.tabset}

```{r}
# Specify timepoint comparisons
timepoints <- paste0("T", 0:3)
tp_pairs <- combn(timepoints, 2, simplify = FALSE)
cmp_corr <- c(
  lapply(tp_pairs, \(p) paste0("R_",  p)),
  lapply(tp_pairs, \(p) paste0("NR_", p)),
  list(
    c("R_T0", "NR_T1"),
    c("NR_T0", "R_T1")
  )
)
```

```{r}
# Calculate correlation of fold changes between different timepoints
cor_mat <- list()
for (cmp in cmp_corr) {

  cmp_lab <- paste(cmp[[1]], "vs", cmp[[2]], sep = "_")

  resp1 <- ifelse(str_starts(cmp[[1]], "NR"), "NR", "R")
  resp2 <- ifelse(str_starts(cmp[[2]], "NR"), "NR", "R")
  tp1 <- str_extract(cmp[[1]], "T[0-9]+")
  tp2 <- str_extract(cmp[[2]], "T[0-9]+")

  ps_1 <- ps %>%
    filter(response == resp1, timepoint == tp1) %>%
    select(patientid, peptide_id, fold_change, exist)

  ps_2 <- ps %>%
    filter(response == resp2, timepoint == tp2) %>%
    select(patientid, peptide_id, fold_change, exist)

  # Fold change matrices (Pearson)
  mat1_fc <- ps_1 %>%
    distinct() %>%
    arrange(patientid) %>%
    collect() %>%
    pivot_wider(
      id_cols = peptide_id, 
      names_from = patientid, 
      values_from = fold_change
    ) %>%
    column_to_rownames("peptide_id") %>%
    as.matrix()

  mat2_fc <- ps_2 %>%
    distinct() %>%
    arrange(patientid) %>%
    collect() %>%
    pivot_wider(
      id_cols = peptide_id, 
      names_from = patientid, 
      values_from = fold_change
    ) %>%
    column_to_rownames("peptide_id") %>%
    as.matrix()

  # Exist matrices (Kulczynski)
  mat1_ex <- ps_1 %>%
    distinct() %>%
    arrange(patientid) %>%
    collect() %>%
    pivot_wider(
      id_cols = peptide_id, 
      names_from = patientid, 
      values_from = exist
    ) %>%
    column_to_rownames("peptide_id") %>%
    as.matrix()

  mat2_ex <- ps_2 %>%
    distinct() %>%
    arrange(patientid) %>%
    collect() %>%
    pivot_wider(
      id_cols = peptide_id, 
      names_from = patientid, 
      values_from = exist
    ) %>%
    column_to_rownames("peptide_id") %>%
    as.matrix()

  # Align peptides across comparisons
  common_peptides <- intersect(rownames(mat1_fc), rownames(mat2_fc))
  mat1_fc <- mat1_fc[common_peptides, ]
  mat2_fc <- mat2_fc[common_peptides, ]
  mat1_ex <- mat1_ex[common_peptides, ]
  mat2_ex <- mat2_ex[common_peptides, ]

  # Compute matrices
  pearson <- cor(mat1_fc, mat2_fc, use = "pairwise.complete.obs")
  kulc <- kulczynski_mat(mat1_ex, mat2_ex)

  # Convert to long table
  pearson_df <- pearson %>%
    as.data.frame() %>%
    rownames_to_column("pid_t1") %>%
    pivot_longer(-pid_t1, names_to = "pid_t2", values_to = "pearson")
  kulc_df <- kulc %>%
    as.data.frame() %>%
    rownames_to_column("pid_t1") %>%
    pivot_longer(-pid_t1, names_to = "pid_t2", values_to = "kulczynski")

  # Combine
  cor_mat[[cmp_lab]] <- left_join(
    pearson_df, 
    kulc_df, 
    by = c("pid_t1", "pid_t2")
  )
}
```

```{r results='asis', fig.width=15, fig.height=6}
# Heatmap: Pearson + Kulczynski (T0/T1)
cmp_heat <- list(c("R_T0", "R_T1"), c("NR_T0", "NR_T1"))
metrics <- c("pearson", "kulczynski")
for (cmp in cmp_heat) {
  
  cmp_lab <- paste(cmp[[1]], "vs", cmp[[2]], sep = "_")
  resp <- ifelse(str_starts(cmp[[1]], "NR"), "NR", "R")

  # Keep only patients occurring in both timepoints
  common_patients <- intersect(
    unique(cor_mat[[cmp_lab]]$pid_t1), 
    unique(cor_mat[[cmp_lab]]$pid_t2)
  )
  common_patients <- sort(common_patients)  # deterministic order

  # Create sequential labels
  new_ids <- as.character(seq_along(common_patients))
  ids_map <- setNames(new_ids, common_patients)

  df_hm <- cor_mat[[cmp_lab]] %>%
    filter(pid_t1 %in% common_patients, pid_t2 %in% common_patients) %>%
    mutate(
      pid_t1_num = factor(ids_map[as.character(pid_t1)], levels = new_ids),
      pid_t2_num = factor(ids_map[as.character(pid_t2)], levels = new_ids)
    )

  # Create a tab header (level below the {.tabset} header)
  cat("### ", cmp_lab, "\n\n", sep = "")

  p_heat <- lapply(metrics, function(metric) {
    
    # Create patient labels
    x_labs <- levels(df_hm$pid_t1_num)
    y_labs <- levels(df_hm$pid_t2_num)
    x_labs[seq_along(x_labs) %% 2 == 0] <- ""
    y_labs[seq_along(y_labs) %% 2 == 0] <- ""
    
    df_hm %>%
      ggplot(aes(x = pid_t1_num, y = pid_t2_num, fill = .data[[metric]])) +
      geom_tile() +
      scale_fill_viridis_c(
        option = "plasma",
        # limits = c(0.15, 1),
        breaks = seq(0.2, 0.8, 0.2),
        # oob = scales::squish,
        guide = guide_colorbar(
          barheight = unit(0.85, "npc"), 
          barwidth  = unit(0.5, "cm")
        )
      ) +
      scale_x_discrete(labels = x_labs) +
      scale_y_discrete(labels = y_labs) +
      coord_fixed() +
      theme(
        text = element_text(family = "Montserrat", size = 12),
        panel.grid = element_blank(),
        legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size = 12)
      ) +
      labs(
        title = paste0("Correlation between ", resp, ": T0 vs. T1 Epitope Repertoires"),
        subtitle = metric,
        x = paste0(resp, " T0 Ig epitope repertoire (individual count/ID)"),
        y = paste0(resp, " T1 Ig epitope repertoire (individual count/ID)")
      )
  })
  pw <- wrap_plots(p_heat)
  print(pw)
  cat("\n\n")
}
```

### Pearson (T0/T1) - R and NR combined

```{r, fig.width=3.5, fig.height=3.5}
# Heatmap: Pearson (T0/T1) - R and NR combined, matched patients only

# Four blocks
df_rr <- cor_mat[["R_T0_vs_R_T1"]]
df_rn <- cor_mat[["R_T0_vs_NR_T1"]]
df_nr <- cor_mat[["NR_T0_vs_R_T1"]]
df_nn <- cor_mat[["NR_T0_vs_NR_T1"]]

# Keep only patients occurring in both timepoints within each response group
common_r <- sort(intersect(unique(df_rr$pid_t1), unique(df_rr$pid_t2)))
common_nr <- sort(intersect(unique(df_nn$pid_t1), unique(df_nn$pid_t2)))

# Create sequential labels
r_ids  <- setNames(as.character(seq_along(common_r)),  common_r)
nr_ids <- setNames(as.character(seq_along(common_nr)), common_nr)

# Filter all four blocks consistently
df_rr_f <- df_rr %>%
  filter(pid_t1 %in% common_r, pid_t2 %in% common_r) %>%
  mutate(
    pid_t1_num = factor(r_ids[pid_t1], levels = as.character(seq_along(common_r))),
    pid_t2_num = factor(r_ids[pid_t2], levels = as.character(seq_along(common_r)))
  )

df_rn_f <- df_rn %>%
  filter(pid_t1 %in% common_r, pid_t2 %in% common_nr) %>%
  mutate(
    pid_t1_num = factor(r_ids[pid_t1], levels = as.character(seq_along(common_r))),
    pid_t2_num = factor(nr_ids[pid_t2], levels = as.character(seq_along(common_nr)))
  )

df_nr_f <- df_nr %>%
  filter(pid_t1 %in% common_nr, pid_t2 %in% common_r) %>%
  mutate(
    pid_t1_num = factor(nr_ids[pid_t1], levels = as.character(seq_along(common_nr))),
    pid_t2_num = factor(r_ids[pid_t2], levels = as.character(seq_along(common_r)))
  )

df_nn_f <- df_nn %>%
  filter(pid_t1 %in% common_nr, pid_t2 %in% common_nr) %>%
  mutate(
    pid_t1_num = factor(nr_ids[pid_t1], levels = as.character(seq_along(common_nr))),
    pid_t2_num = factor(nr_ids[pid_t2], levels = as.character(seq_along(common_nr)))
  )

# Axis ordering
r_x_ids  <- levels(df_rr_f$pid_t1_num)
nr_x_ids <- levels(df_nn_f$pid_t1_num)
r_y_ids  <- levels(df_rr_f$pid_t2_num)
nr_y_ids <- levels(df_nn_f$pid_t2_num)

x_levels <- c(paste0("R_", r_x_ids), paste0("NR_", nr_x_ids))
y_levels <- c(paste0("R_", r_y_ids), paste0("NR_", nr_y_ids))

x_labels <- c(
  setNames(ifelse(seq_along(r_x_ids) %in% c(1, 10, 20, 30, 40), seq_along(r_x_ids), ""),
           paste0("R_", r_x_ids)),
  setNames(ifelse(seq_along(nr_x_ids) %in% c(1, 10, 20, 30, 40), seq_along(nr_x_ids), ""),
           paste0("NR_", nr_x_ids))
)

y_labels <- c(
  setNames(ifelse(seq_along(r_y_ids) %in% c(1, 10, 20, 30, 40), seq_along(r_y_ids), ""),
           paste0("R_", r_y_ids)),
  setNames(ifelse(seq_along(r_y_ids) %in% c(1, 10, 20, 30, 40), seq_along(nr_y_ids), ""),
           paste0("NR_", nr_y_ids))
)

# Map each filtered block onto the combined axes
df_comb <- bind_rows(
  df_rr_f %>% mutate(
    x_comb = paste0("R_", pid_t1_num),
    y_comb = paste0("R_", pid_t2_num)
  ),
  df_rn_f %>% mutate(
    x_comb = paste0("R_", pid_t1_num),
    y_comb = paste0("NR_", pid_t2_num)
  ),
  df_nr_f %>% mutate(
    x_comb = paste0("NR_", pid_t1_num),
    y_comb = paste0("R_", pid_t2_num)
  ),
  df_nn_f %>% mutate(
    x_comb = paste0("NR_", pid_t1_num),
    y_comb = paste0("NR_", pid_t2_num)
  )
) %>%
  mutate(
    x_comb = factor(x_comb, levels = x_levels),
    y_comb = factor(y_comb, levels = y_levels)
  )

p_heat_comb <- ggplot(df_comb, aes(x = x_comb, y = y_comb, fill = pearson)) +
  geom_tile() +
  scale_fill_viridis_c(
    option = "plasma",
    breaks = seq(0, 1, 0.1),
    guide = guide_colorbar(
      barheight = unit(0.75, "npc"),
      barwidth  = unit(0.3, "cm"),
      frame.colour = "black",
      frame.linewidth = 0.2
    )
  ) +
  scale_x_discrete(
    breaks = c(
      paste0("R_", r_x_ids[c(1, seq(5, length(r_x_ids), 5))]),
      paste0("NR_", nr_x_ids[c(1, seq(5, length(nr_x_ids), 5))])
    ),
    labels = c(
      x_labels[paste0("R_", r_x_ids[c(1, seq(5, length(r_x_ids), 5))])],
      x_labels[paste0("NR_", nr_x_ids[c(1, seq(5, length(nr_x_ids), 5))])]
    )
  ) +
  scale_y_discrete(
    breaks = c(
      paste0("R_", r_y_ids[c(1, seq(5, length(r_y_ids), 5))]),
      paste0("NR_", nr_y_ids[c(1, seq(5, length(nr_y_ids), 5))])
    ),
    labels = c(
      y_labels[paste0("R_", r_y_ids[c(1, seq(5, length(r_y_ids), 5))])],
      y_labels[paste0("NR_", nr_y_ids[c(1, seq(5, length(nr_y_ids), 5))])]
    )
  ) +
  coord_fixed() +
  geom_vline(
    xintercept = length(common_r) + 0.5,
    linewidth = 0.35, 
    colour = "white", 
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = length(common_r) + 0.5,
    linewidth = 0.35, 
    colour = "white", 
    linetype = "dashed"
  ) +
  theme_bw(base_family = "Arial", base_size = 10) +
  theme(
    plot.title = element_text(size = 12),
    plot.subtitle = element_text(size = 10),
    axis.title = element_blank(),
    axis.text = element_text(size = 10, colour = "black"),
    axis.ticks = element_line(colour = "black", linewidth = 0.3),
    axis.ticks.length = unit(1.5, "mm"),
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    legend.margin = margin(0.5, 0.5, 0.5, 0.5),
    legend.box.margin = margin(0, 0, 0, -5),
    panel.grid = element_blank(),
    plot.margin = margin(5.5, 5.5, 5.5, 5.5, unit = "pt")
  ) +
  labs(
    title = "T0 vs. T1 Ig epitope repertoires",
    subtitle = "Pearson correlation of log(fold change)"
  )

print(p_heat_comb)
```

## Heatmap - Outlier {.tabset}

I wanted to investigate the outlier in the responder heatmap because it doesn't
make sense to me. First I thought maybe those two samples both have a very low
number of enriched peptides, but that's not the case. Then I thought maybe the
sample labels were swapped, but then the correlation should be symmetric. I now
have the suspicion that patient 23 (T1) wasn't measured and is actually a sample 
from patient PBT313 that was measured twice?

* Patient 23 (T0) doesn't correlate with anything
* For patient PBT313 timepoints correlate with each other
* And they both correlate with patient 23 (T1)

```{r}
checks <- c("T0_PBT313", "T1_PBT313", "T0_23", "T1_23")
hmout_corr <- ps %>%
  mutate(label = paste(timepoint, patientid, sep = "_")) %>%
  filter(label %in% checks) %>%
  select(peptide_id, fold_change, label) %>%
  collect() %>%
  pivot_wider(names_from = "label", values_from = "fold_change") %>% 
  select(all_of(checks))

ggpairs(
  hmout_corr,
  upper = list(continuous = wrap("cor", size = 4, use = "pairwise.complete.obs")),
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.5)),
  diag  = list(continuous = wrap("blankDiag"))
) +
  theme_bw()
```

## Correlation coefficients (Histogram) {.tabset}

```{r}
# Extract correlations of matched/unmatched samples
cor_stats <- list()
for (cmp in names(cor_mat)[1:12]) {

  # Keep only patients occurring in both timepoints
  common_patients <- intersect(
    unique(cor_mat[[cmp]]$pid_t1), 
    unique(cor_mat[[cmp]]$pid_t2)
  )

  matched <- cor_mat[[cmp]] %>%
    filter(pid_t1 == pid_t2) %>%
    pull(pearson)
  
  unmatched <- cor_mat[[cmp]] %>%
    filter(pid_t1 %in% common_patients, pid_t2 %in% common_patients) %>%
    filter(pid_t1 != pid_t2) %>%
    pull(pearson)
  
  cor_stats[[cmp]] <- list(matched = matched, unmatched = unmatched)
}
```

```{r}
plot_um_m_corr <- function(cor_um,
                           cor_m,
                           y_lim_up = 200,
                           scale_fct = 10,
                           group = "Group",
                           bins = 40,
                           col_pal = c(um = "grey70", m = "steelblue")) {
  
  df_um <- data.frame(r = cor_um, type = "um")
  df_m <- data.frame(r = cor_m, type = "m")
  
  ggplot() +
    geom_histogram(
      data = df_um,
      aes(x = r, y = after_stat(count), fill = type),
      bins = bins,
      alpha = 0.8
    ) +
    geom_histogram(
      data = df_m,
      aes(x = r, y = after_stat(count) * scale_fct, fill = type),
      bins = bins,
      alpha = 0.8
    ) +
    scale_y_continuous(
      limits = c(0, y_lim_up),
      name = "Unmatched individuals sample correlation",
      sec.axis = sec_axis(
        ~ . / scale_fct, 
        name = "Matched individuals sample correlation"
      )
    ) +
    scale_fill_manual(values = col_pal) +
    labs(
      subtitle = group,
      x = "Pearson correlation of \nlog(fold change) T0 vs. T1"
    ) +
    theme_bw(base_size = 12) +
    theme(
      text = element_text(family = "Montserrat"),
      axis.title.y.right = element_text(margin = margin(l = 8), angle = 90, vjust = 0.5),
      axis.title.y.left  = element_text(margin = margin(r = 8)),
      panel.grid = element_blank(),
      legend.position = "none"
    )
}
```

```{r, fig.width=7, fig.height=3.5}
p1 <- plot_um_m_corr(
  cor_stats$R_T0_vs_R_T1$unmatched,
  cor_stats$R_T0_vs_R_T1$matched,
  y_lim_up = 180,
  scale_fct = 10,
  bins = 40,
  group = "Responders",
  col_pal = c(um = "grey70", m = group_palette[["R"]])
)

p2 <- plot_um_m_corr(
  cor_stats$NR_T0_vs_NR_T1$unmatched,
  cor_stats$NR_T0_vs_NR_T1$matched,
  y_lim_up = 180,
  scale_fct = 10,
  bins = 40,
  group = "Non-responders",
  col_pal = c(um = "grey70", m = group_palette[["NR"]])
)

wrap_plots(p1, p2, ncol = 2)
```

## Correlation coefficients (Barplots) {.tabset}

- Do we want statistical significance here? They are all non-significant.
- Add the number of samples that were considered in each comparison?

```{r}
# Extract just the correlations of the matched samples
matched_df <- bind_rows(
  lapply(names(cor_stats), function(nm) {
    vals <- cor_stats[[nm]]$matched
    data.frame(contrast = nm, value = vals)
  })
) %>%
  mutate(
    group = str_extract(contrast, "(R|NR)"),
    time_comp = str_match(contrast, "T(\\d+)_vs_(?:NR|R)_T(\\d+)") |> 
      (\(m) paste0("T", m[,2], "/T", m[,3]))()  
    )

# Compute significance between R/NR at each timepoint comparison
stats_tests <- compare_means(
  value ~ group,
  data  = matched_df,
  group.by = "time_comp",
  method = "wilcox.test",
  p.adjust.method = "bonferroni"
) %>%
  rstatix::add_significance("p.adj")

# Put everything together
matched_df <- matched_df %>%
  group_by(time_comp, group) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    sd   = sd(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(stats_tests, by = "time_comp") %>%
  mutate(group = factor(group, levels = c("R", "NR")))
```

```{r, fig.width=5, fig.height=5}
ggplot(matched_df, aes(x = time_comp, y = mean, fill = group)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(
    aes(ymin = mean - sd, ymax = mean + sd),
    position = position_dodge(width = 0.7),
    width = 0.2
  ) +
  geom_text(
    aes(x = time_comp, y = 1.1, label = p.adj.signif),
    inherit.aes = FALSE,
    vjust = 0
  ) +
  scale_fill_manual(values = group_palette) +
  labs(
    x = NULL,
    y = "Mean Pearson correlation (matched samples)"
  ) +
  theme_bw(12) +
  theme(
    panel.grid = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```


# Significant Taxa / Annotations {.tabset}

## POP framework {.tabset}

```{r}
pop_df %>%
  filter(category_rank_bh != "not significant") %>%
  arrange(p_adj_rank, p_raw) %>%
  relocate(p_raw, p_adj_rank, percent1, percent2, group1, group2, .after = feature) %>%
  head(10) %>%
  gt()
```

## DELTA framework {.tabset}

```{r}
delta_df %>%
  filter(category_rank_bh != "not significant") %>%
  arrange(p_adj_rank, p_perm) %>%
  relocate(p_adj_rank, p_perm, Z_from_p, .after = feature) %>%
  head(10) %>%
  gt()
```


# Figure 2 {.tabset}

```{r}
# Set colours
peptide_palette <- c(
  "Complete library"   = "#9467bd",
  "Microbiome"         = "#ff9896",
  "Pathogenic strains" = "#d62728",
  "IgA"                = "#8c564b",
  "Probiotic"          = "#98df8a",
  "Flagellins"         = "#2ca02c",
  "VFDB"               = "#e377c2",
  "Phages"             = "#bcbd22",
  "Allergens"          = "#17becf",
  "IEDB controls"      = "#7f7f7f"
)
```

```{r}
make_scatter_flag <- function(pep_tbl, pval_mw_df, phip_palette) {
  # Apply filter
  pep_tbl_prev10 <- pep_tbl %>% filter((prop1 + prop2 >= 0.1))
  
  # Get peptide library
  peptide_library <- attr(pep_tbl, "prev_meta")$peptide_library

  N1 <- unique(pep_tbl_prev10$N1)
  N2 <- unique(pep_tbl_prev10$N2)

  ## Flagellins scatter plot
  # Extract adjusted p-value
  pval_mw <- pval_mw_df %>% 
    filter(group1 == "Complete library", group2 == "Flagellins") %>%
    pull(p.adj)
  
  # Removing datapoints where is_flagellum is NA, so NAs won't be plotted
  # Alternatively could set NAs to FALSE in peptide_library to plot NAs too
  keep_flag <- peptide_library %>% filter(!is.na(is_bac_flagella))
  pep_tbl_flag <- pep_tbl_prev10 %>% filter(feature %in% keep_flag$peptide_id)
  
  legend_lab <- paste0("Flagellins (p = ", format_pval(pval_mw), ")")
  
  p_static <- scatter_static(
    df = pep_tbl_flag,
    rank = "peptide_id",
    xlab = unique(pep_tbl$group1),
    ylab = unique(pep_tbl$group2),
    color_by = "is_bac_flagella",
    color_title = NULL,
    point_size = 2,
    point_alpha = 0,
    font_size = 12
  ) +
    coord_cartesian(xlim = c(-2, 102), ylim = c(-2, 102), expand = TRUE) +
    theme(
      text = element_text(size = 12, family = "Montserrat"),
      legend.position = c(0, 1), 
      legend.justification = c("left", "top"), 
      legend.title = element_blank(),
      axis.title = element_text(face = "plain"),
      legend.text = element_text(size = 12),
      plot.margin = margin(5.5, 5.5, 5.5, 5.5, unit = "pt")
    ) +
    labs(
      x = paste0(
        "% of ", unique(pep_tbl$group1), " with a peptide \nsignificantly bound ",
        "(n = ", N1, ")"
      ),
      y = paste0(
        "% of ", unique(pep_tbl$group2), " with a peptide \nsignificantly bound ",
        "(n = ", N2, ")"
      )
    )
  
  p_static$data <- p_static$data %>%
    mutate(
      pt_size = if_else(is_bac_flagella == 1, 2, 1),
      legend_group = if_else(is_bac_flagella == 1, legend_lab, NA_character_)
    )

  p_static <- p_static +
    geom_point(
      data = p_static$data %>% filter(is_bac_flagella != 1),
      aes(x = percent1, y = percent2),
      colour = "grey70",
      size = 1.2,
      alpha = 0.5,
      inherit.aes = FALSE,
      show.legend = FALSE
    ) +
    geom_point(
      data = p_static$data %>% filter(is_bac_flagella == 1),
      aes(x = percent1, y = percent2, colour = legend_group),
      size = 2,
      alpha = 0.5,
      inherit.aes = FALSE
    ) +
    scale_colour_manual(
      values = c(setNames(peptide_palette[["Flagellins"]], legend_lab))
    ) +
    scale_size_identity() +
    guides(colour = guide_legend(override.aes = list(size = 2.5, alpha = 0.5)))

  return(p_static)
}
```

## Peptide Comparison Between Groups (POP) {.tabset}

```{r results='asis', fig.width=5, fig.height=5}
cmp_scat <- c("R_T0_vs_NR_T0", "R_T1_vs_NR_T1", "R_T0_vs_R_T1", "NR_T0_vs_NR_T1")
for (nm in cmp_scat) {
  
  # Create a tab header (level below the {.tabset} header)
  cat("### ", nm, "\n\n", sep = "")
  
  p <- make_scatter_flag(
    pep_tbl = pep_tbl[[nm]],
    pval_mw_df = pval_mw_df[[nm]],
    phip_palette = phip_palette
  )
  
  print(p)
  cat("\n\n")
}
```

## Ratios of response by oligo family {.tabset}

- I set an arbitrary prevalence threshold of at least prop1 + prop2 >= 0.1
- For some of the library categories not sure if they are correct?
- Significance calulated with Wilcoxon and Bonferroni corrected
- Can't seem to reproduce the same plot. What test?
- How are ratios calculated (to avoid Inf)?

```{r}
# Apply filter
pep_tbl_prev10 <- pep_tbl$R_T0_vs_NR_T0 %>% filter((prop1 + prop2 >= 0.1))

# Add peptide library
pep_tbl_peplib <- pep_tbl_prev10 %>%
  left_join(
    peplib,
    by = c("feature" = "peptide_id")
  ) %>%
  mutate(is_library = 1)

N1 <- unique(pep_tbl_peplib$N1)
N2 <- unique(pep_tbl_peplib$N2)

# Prepare long format for boxplots by annotation
family_cols <- c(
  "is_library"       = "Complete library",
  "is_MPA"           = "Microbiome",
  "is_patho"         = "Pathogenic strains",
  "is_IgA"           = "IgA",
  "is_probio"        = "Probiotic",
  "is_bac_flagella"  = "Flagellins",
  "is_infect"        = "VFDB",
  "is_phage"         = "Phages",
  "is_allergens"     = "Allergens",
  "is_IEDB_or_cntrl" = "IEDB controls"
)

pep_tbl_peplib_long <- pep_tbl_peplib %>%
  select(feature, ratio, all_of(names(family_cols))) %>%
  pivot_longer(
    cols = all_of(names(family_cols)),
    names_to = "family",
    values_to = "present"
  ) %>%
  filter(present == 1) %>%
  mutate(family_lab = factor(family_cols[family], levels = family_cols))
```

```{r, fig.width=10, fig.height=5}
## Boxplot - peptide categories (each dot = peptide)
# Remove outliers for plotting (nicer plots)
pep_tbl_peplib_long <- pep_tbl_peplib_long %>%
  group_by(family_lab) %>%
  filter(!ratio %in% boxplot.stats(ratio)$out)

# Stacked y positions so brackets don't overlap
max_y <- max(pep_tbl_peplib_long$ratio, na.rm = TRUE)

# Keep factor order exactly as used in the plot
x_levels <- levels(factor(pep_tbl_peplib_long$family_lab))
flag_pos <- match("Flagellins", x_levels)

# Build p-value table for only comparisons involving flagellins
pdat <- pval_mw_df$R_T0_vs_NR_T0 %>%
  filter(group1 == "Flagellins" | group2 == "Flagellins") %>%
  mutate(
    other_group = if_else(group1 == "Flagellins", group2, group1),
    other_pos = match(other_group, x_levels),
    side = if_else(other_pos < flag_pos, "left", "right"),
    dist = abs(other_pos - flag_pos)
  ) %>%
  # Make sure xmin is always the left category and xmax the right category
  mutate(
    xmin = if_else(other_pos < flag_pos, other_group, "Flagellins"),
    xmax = if_else(other_pos < flag_pos, "Flagellins", other_group)
  ) %>%
  # Alternate left/right from inside to outside
  arrange(dist, factor(side, levels = c("left", "right"))) %>%
  mutate(
    y.position = max_y * (1 + 0.11 * row_number())
  ) %>%
  rstatix::add_significance("p.adj")

# Boxplot (each dot = peptide)
p_pep_box <- ggplot(
  pep_tbl_peplib_long, 
  aes(x = family_lab, y = ratio, fill = family_lab)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  scale_fill_manual(values = peptide_palette) +
  stat_pvalue_manual(
    pdat,
    label = "p.adj.signif",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.02,
    size = 3.51
  ) +
  coord_cartesian(ylim = c(0, 6)) +
  labs(
    x = NULL, 
    y = paste0(
      "Ratios of Ig response in \nR_T0_vs_NR_T0",
      " (", N1, ", ", N2, ")"
    )
  ) +
  theme_phip(12) +
  theme(
    text = element_text(size = 12, family = "Montserrat"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title = element_text(face = "plain"),
    plot.margin = margin(5.5, 5.5, 5.5, 5.5, unit = "pt")
  )

print(p_pep_box)
```

## Ratios of response to flagellins by class at different timepoints {.tabset}

```{r}
flagellins <- pep_tbl_peplib %>% filter(is_bac_flagella == 1) %>% pull(feature)
class_lvls <- c("response (R/NR)", unique(labs$group_type))

pep_tbl_all <- c(pep_tbl[1:4], pep_tbl_xtra)
pval_mw_df_all <- c(pval_mw_df[1:4], pval_mw_df_xtra)

class_tp_df <- bind_rows(pep_tbl_all, .id = "contrast") %>%
  filter(feature %in% flagellins) %>%
  mutate(
    tp = str_extract(group1, "T\\d+$"),
    grp = str_remove(group1, "_T\\d+$"),
    class = case_when(
      grp %in% c("R", "NR") ~ "response (R/NR)",
      str_detect(grp, "_") ~ str_extract(grp, "^[^_]+"),
      TRUE ~ grp
    ),
    class = factor(class, levels = class_lvls)
  )

# Remove outliers for plotting (nicer plots)
class_tp_df <- class_tp_df %>%
  group_by(class, tp) %>%
  filter(!ratio %in% boxplot.stats(ratio)$out)

# Calculate the top of the whisker
ypos_df <- class_tp_df %>%
  group_by(class, tp) %>%
  summarise(
    y.position = boxplot.stats(ratio)$stats[5] + 0.5, 
    .groups = "drop"
  )

# Build p-value table for flagellins vs. complete library only
pdat <- bind_rows(pval_mw_df_all, .id = "contrast") %>%
  filter(
    (group1 == "Flagellins" & group2 == "Complete library") | 
    (group1 == "Complete library" & group2 == "Flagellins")
  ) %>%
  mutate(p.adj.signif = if_else(p.adj.signif == "ns", "", p.adj.signif)) %>%
  select(contrast, p.adj.signif) %>%
  left_join(
    class_tp_df %>% distinct(contrast, group1, group2, class, tp),
    by = "contrast"
  ) %>%
  left_join(ypos_df, by = c("class", "tp"))
```

```{r, fig.width=9, fig.height=4.5}
p_class_tp_box <- class_tp_df %>%
  ggplot(aes(x = class, y = ratio, fill = tp)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(
    x = NULL,
    y = "Ratios of Ig response to \nflagellins by class (yes/no)", 
    fill = "Timepoint"
  ) +
  scale_fill_manual(values = c(
    "T0" = "grey70",
    "T1" = "#6BBF58",
    "T2" = "#59A14F",
    "T3" = "#3D702D"
  )) +
  geom_text(
    data = pdat,
    aes(x = class, y = y.position, label = p.adj.signif, group = tp),
    position = position_dodge(width = 0.8),
    inherit.aes = FALSE,
    size = 3.51,
    angle = 90,
    hjust = 0,
    vjust = 0.7
  ) +
  theme_phip(12) +
  theme(
    text = element_text(size = 12, family = "Montserrat"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title = element_text(face = "plain"),
    plot.margin = margin(5.5, 5.5, 5.5, 5.5, unit = "pt"),
    legend.position = "right",
    legend.direction = "vertical",
    legend.title = element_text(face = "plain")
  )

print(p_class_tp_box)
```


# Additional Plots

- Here the dates are way more extreme than in the original figure. Why?
- Used serum_collection_date-start_io (weeks)

```{r}
meta %>%
  mutate(
    baseline_date = as.Date(start_io, "%d/%m/%Y"),
    timepoint_date = as.Date(serum_collection_date, "%d/%m/%Y"),
    diff_tp_base_week = difftime(timepoint_date, baseline_date, units = "weeks")
  ) %>%
  ggplot(aes(x = timepoint, y = diff_tp_base_week)) +
  geom_boxplot(outliers = TRUE)
```


# Survival analysis {.tabset}

The survival data only has sample IDs (stool samples), no patient IDs. However,
I was able to uniquely match patients based on age, sex and BMI.

```{r}
print("Patients with PhIP-Seq data (T1):")
meta %>% filter(timepoint == "T1") %>% nrow()

print("Patients with PhIP-Seq data (T1) and survival data (initial metadata):")
with_surv <- meta %>%
  filter(timepoint == "T1") %>%
  mutate(bmi = round(bmi, 3)) %>%
  left_join(
    meta_surv %>% mutate(bmi = round(bmi, 3)), 
    by = c("sex", "age", "bmi", "centre")
  ) %>% 
  filter(!is.na(os_months)) %>%
  select(LV_code, patientid, centre, response, os_months)

with_surv %>% nrow()

print("Patients with PhIP-Seq data (T1) and survival data (new metadata):")
with_surv_new <- meta %>%
  filter(timepoint == "T1") %>%
  left_join(
    meta_surv_new, 
    by = c("patientid")
  ) %>% 
  filter(!is.na(time_months)) %>%
  select(LV_code, patientid, centre, response, time_months)

with_surv_new %>% nrow()
```

```{r}
with_surv_comb <- with_surv %>% full_join(with_surv_new) %>% distinct()

with_surv_comb %>% nrow()
table(with_surv_comb$response)
table(with_surv_comb$centre)

# One patient with different OS depending on metadata (LV1016273389)
```
