---
title: "Figures for PRIMM PhIP-Seq paper"
subtitle: "Plots created with phiper package"
author: "Magdalena Hübner"
license: "CC-BY 4.0"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    theme: united
    highlight: tango
    toc: true              # Enable table of contents
    toc_float: true        # Make the table of contents float on the side
    number_sections: true  # Optional: Number your sections
    toc_depth: 4  
---

<!-- R markdown settings -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE,
  fig.width = 5,
  fig.height = 4
)
```

<!-- Project setup -->

```{r}
# Core tidyverse
library(purrr)
library(stringr)
library(tidyr)
library(tibble)
library(dplyr)

# I/O
library(here)
library(openxlsx)

# Plotting
library(GGally)
library(ggplot2)
library(ggpubr)
library(gt)
library(patchwork)

# PhIP-Seq
library(phiper)
```

```{r}
# Seed (for jitter etc.)
set.seed(612)

# Suppress phiper logging in report
options(phiper.log.verbose = FALSE)
```

```{r}
# Colours scheme for plotting
group_palette <- c(
  # Responders
  "R" = phip_palette[1],
  "R_T1" = phip_palette[1],
  "R_T2" = phip_palette[1],
  # Non-responders
  "NR" = phip_palette[9],
  "NR_T1" = phip_palette[9],
  "NR_T2" = phip_palette[9]
)
```

<!-- Loading functions and pre-computed data -->

```{r}
source(here("R", "helpers.R"))
```

```{r}
# Path to original data from Israel
zip_path <- "/lisc/data/work/ccr/SHARED_RESOURCES/external_data/PRIMM/FinalB_2.zip"
```

```{r}
# PhIP-Seq data
ps <- phip_convert(here("data/processed/", "primm_full.parquet")) %>%
  mutate(group_char = paste(response, timepoint, sep = "_"))

# Sample metadata
meta <- read.csv(here("data/processed/", "metadata.csv")) %>%
  mutate(group_char = paste(response, timepoint, sep = "_"))

# Peptide library
peplib <- readRDS(here("results", "peptide_library.rds"))

# Alpha diversity
alpha_div <- read.xlsx(here("results/R_T1_vs_NR_T1", "alpha_diversity", "table.xlsx"))

# Beta diversity
pcoa_res <- list()
perm_res <- list()
disp_res <- list()
for(cmp in c("R_T1_vs_NR_T1", "R_T2_vs_NR_T2")) {
  pcoa_res[[cmp]] <- readRDS(here("results", cmp, "beta_diversity", "pcoa_results.rds"))
  perm_res[[cmp]] <- readRDS(here("results", cmp, "beta_diversity", "permanova_results.rds"))
  disp_res[[cmp]] <- readRDS(here("results", cmp, "beta_diversity", "dispersion_results.rds"))
}

# POP framework
pep_tbl <- list()
for(cmp in c("R_T1_vs_NR_T1", "R_T2_vs_NR_T2", "R_T1_vs_R_T2", "NR_T1_vs_NR_T2")) {
  pep_tbl[[cmp]] <- readRDS(here("results", cmp, "POP_framework", "single_peptide.rds"))
}
pop_df <- read.csv(
  here("results/R_T1_vs_NR_T1/POP_framework/taxa_ranks.csv"),
  row.names = 1
)

# DELTA framework
delta_df <- read.csv(
  here("results/R_T1_vs_NR_T1/DELTA_framework/delta_table.csv"),
  row.names = 1
)
```

```{r}
# Survival data
meta_surv <- read.csv(here("data/raw/mdat_baseline_survival.csv"))
```

```{r}
# POP framework (other contrasts)
# Vector of group variables and timepoints
group_vars <- c("combiIO", "PFS12", "ORR", "toxicity", "colitis", "antibiotics", "ppi")
timepoints <- c("T1","T2","T3","T4")
levels2 <- c("yes","no")

# All group labels that to consider
labs <- expand_grid(
  group_type = group_vars,
  group_value = levels2,
  timepoint = timepoints
) %>%
  mutate(
    lab = str_c(group_type, group_value, timepoint, sep = "_"),
    group_lvl = str_c(group_type, group_value, sep = "_")
  )

# 1) within-timepoint: yes vs no
cmp_within <- labs %>%
  select(group_type, group_value, timepoint, lab) %>%
  pivot_wider(names_from = group_value, values_from = lab) %>%
  transmute(cmp = map2(yes, no, c))

# 2) across-timepoint within level: yes_Ti vs yes_Tj AND no_Ti vs no_Tj
cmp_across <- labs %>%
  group_by(group_type, group_value) %>%
  summarise(cmp = list(combn(lab, 2, simplify = FALSE)), .groups = "drop") %>%
  unnest(cmp)

comp_xtra <- c(cmp_within$cmp, cmp_across$cmp)
pep_tbl_xtra <- list()
for(cmp_pair in comp_xtra) {
  cmp <- paste0(cmp_pair[1], "_vs_", cmp_pair[2])
  pep_tbl_xtra[[cmp]] <- read.csv(here("results", cmp, "POP_framework", "single_peptide.csv"))
}
```


# Metadata

I noticed that there is one sample (**LV1016273557**) which has two metadata
entries belonging to two completely different patients. 

```{r}
meta_all <- read.csv(here("data/processed/", "metadata_all.csv")) %>%
  mutate(group_char = paste(response, timepoint, sep = "_")) %>%
  select(LV_code, patientid, timepoint, centre, age, Sample.ID, response)

meta_all %>% filter(LV_code == "LV1016273557")
meta_all %>% filter(patientid == "MVCC011")
meta_all %>% filter(patientid == "OL_3049")
```

Initially I removed this sample from the analysis.
This is how the plots for Johannes grant were generated.


# Figure 1 {.tabset}

## Alpha Diversity {.tabset}

- Richness: The absolute number of antibody-bound peptides/taxa rank per 
individual between groups.

- Shannon: A given taxonomic rank diversity of Ig repertoires of antibody-bound 
  peptides per individual between groups.
  
**This was done with phiper**

We could consider using a richness cutoff (e.g. 200)?

```{r, fig.width=8, fig.height=4}
# Alpha diversity plot (Richness + Shannon)
metrics <- c("richness", "shannon_diversity")
cmp_alpha <- c("R_T1", "NR_T1")

p_alpha <- lapply(metrics, function(m) {
  # Calculate and format p-values first
  pv <- compare_means(
    formula = reformulate("group_char", response = m),
    data = alpha_div,
    method = "wilcox.test",
    comparisons = list(cmp_alpha)
  ) %>%
    mutate(
      p = format_pval(p),
      y.position = max(alpha_div[[m]], na.rm = TRUE) * 1.1
    )
  
  p_alpha <- phiper::plot_alpha_diversity(
    alpha_div,
    metric = m,
    group_col = "group_char",
    x_order = cmp_alpha
  ) +
    stat_pvalue_manual(
      pv,
      label = "p",
      y.position = "y.position",
      tip.length = 0.02,
      label.size = 4,
      family = "Montserrat"
    ) +
    scale_fill_manual(values = group_palette) +
    coord_cartesian(clip = "off") + 
    theme(
      text = element_text(size = 12),
      axis.title.x = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
    )
  
  if (m == "richness") {
    p_alpha <- p_alpha +
      ggrepel::geom_text_repel(
        data = alpha_div %>% filter(richness<200),
        mapping = aes(label=sample_id),
        size = 3,
        show.legend = FALSE
      )
  }
  p_alpha
})

wrap_plots(p_alpha)
```

## Beta Diversity {.tabset}

- For each peptide Bray–Curtis dissimilarity is computed on the binary abundance 
taxonomic matrix and carried out a Principal Coordinates Analysis (PCoA/MDS). 

- On the scatterplot, each point is a sample and the 95% confidence ellipses 
show the group clustering.

- Tested for overall differences in community composition between groups with 
PERMANOVA, reporting the p-value in the plot. To check that any PERMANOVA result 
wasn’t driven by unequal dispersion, a permutation test on the multivariate 
dispersions was run.

- Together, these panels tell you whether (a) the two groups occupy distinct 
regions of Bray–Curtis space (significant PERMANOVA) and (b) whether one group 
is more dispersed than the other (significant dispersion test).

**This was done with phiper**

I can see 2 clusters at TP1, is this an experimental batch effect?

```{r, fig.width=8, fig.height=4}
# PCoA plots
cmp_pcoa <- list(c("R_T1", "NR_T1"), c("R_T2", "NR_T2"))

p_pcoa <- lapply(cmp_pcoa, function(cmp) {
  # Add group information to PCoA sample coordinates
  cmp_lab <- paste(cmp[[1]], "vs", cmp[[2]], sep = "_")
  pcoa_res[[cmp_lab]]$sample_coords <- pcoa_res[[cmp_lab]]$sample_coords %>%
    left_join(
      meta %>% select(LV_code, group_char),
      by = c("sample_id" = "LV_code")
    ) %>%
    mutate(group_char = factor(group_char, levels = cmp)) %>%
    left_join(
      alpha_div %>% select(sample_id, richness),
      by = "sample_id"
    )
  
  lab_perm <- paste0("PERMANOVA p = ", format_pval(perm_res[[cmp_lab]]$p_adjust))
  lab_disp <- paste0( "Dispersion   p = ", format_pval(disp_res[[cmp_lab]]$tests$p_adjust))

  # PCoA plot with group centroids and ellipses
  phiper:::plot_pcoa(
    pcoa_res[[cmp_lab]],
    axes = c(1, 2),
    group_col = "group_char",
    ellipse_by = "group",
    show_centroids = TRUE,
    point_size = 2
  ) +
    scale_colour_manual(values = group_palette) +
    theme(
      text = element_text(size = 12),
      legend.title = element_blank()
      # legend.position = "none"
    ) +
    ggrepel::geom_text_repel(
      data = pcoa_res[[cmp_lab]]$sample_coords %>% filter(richness < 200),
      mapping = aes(label=sample_id),
      size = 3,
      show.legend = FALSE
    ) +
    annotate(
      "text",
      x = Inf,
      y = -Inf,
      label = paste(lab_perm, lab_disp, sep = "\n"),
      hjust = 1,  # nudge left a bit
      vjust = -0.2,  # nudge up a bit
      size = 4
    )
})

wrap_plots(p_pcoa)
```

## Heatmap {.tabset}

```{r}
# Specify timepoint comparisons
timepoints <- paste0("T", 1:4)
tp_pairs <- combn(timepoints, 2, simplify = FALSE)
cmp_corr <- c(
  lapply(tp_pairs, \(p) paste0("R_",  p)),
  lapply(tp_pairs, \(p) paste0("NR_", p))
)
```

```{r}
# Calculate correlation of fold changes between different timepoints
cor_mat <- list()
for (cmp in cmp_corr) {
  
  cmp_lab <- paste(cmp[[1]], "vs", cmp[[2]], sep = "_")
  resp <- ifelse(str_starts(cmp[[1]], "NR"), "NR", "R")
  tps <- str_extract(cmp, "T[0-9]+")

  ps_sub <- ps %>%
    filter(response == resp, timepoint %in% tps) %>%
    select(sample_id, patientid, timepoint, peptide_id, fold_change, exist)
  
  # Fold change matrices (Pearson)
  mats_fc <- map(tps, function(tp) {
    ps_sub %>%
      filter(timepoint == tp) %>%
      select(peptide_id, patientid, fold_change) %>%
      distinct() %>%
      arrange(patientid) %>%
      pivot_wider(names_from = patientid, values_from = fold_change) %>%
      collect() %>%
      column_to_rownames("peptide_id") %>%
      as.matrix()
  })
  names(mats_fc) <- tps

  # Exist matrices (Kulczynski)
  mats_ex <- map(tps, function(tp) {
    ps_sub %>%
      filter(timepoint == tp) %>%
      select(peptide_id, patientid, exist) %>%
      distinct() %>%
      arrange(patientid) %>%
      pivot_wider(names_from = patientid, values_from = exist) %>%
      collect() %>%
      column_to_rownames("peptide_id") %>%
      as.matrix()
  })
  names(mats_ex) <- tps

  # Align patients and peptides across both metrics
  common_patients <- intersect(colnames(mats_fc[[1]]), colnames(mats_fc[[2]]))
  common_patients <- sort(common_patients) # deterministic order
  mats_fc <- map(mats_fc, ~ .x[, common_patients, drop = FALSE])
  mats_ex <- map(mats_ex, ~ .x[, common_patients, drop = FALSE])

  common_peptides <- intersect(rownames(mats_fc[[1]]), rownames(mats_fc[[2]]))
  mats_fc <- map(mats_fc, ~ .x[common_peptides, , drop = FALSE])
  mats_ex <- map(mats_ex, ~ .x[common_peptides, , drop = FALSE])

  # Compute matrices
  pearson <- cor(mats_fc[[1]], mats_fc[[2]], use = "pairwise.complete.obs")
  kulc <- kulczynski_mat(mats_ex[[1]], mats_ex[[2]])
  
  # Convert to long table
  pearson_df <- pearson %>%
    as.data.frame() %>%
    rownames_to_column("pid_t1") %>%
    pivot_longer(-pid_t1, names_to = "pid_t2", values_to = "pearson")
  kulc_df <- kulc %>%
    as.data.frame() %>%
    rownames_to_column("pid_t1") %>%
    pivot_longer(-pid_t1, names_to = "pid_t2", values_to = "kulczynski")

  # Create sequential labels
  new_ids <- as.character(seq_along(common_patients))
  ids_map <- setNames(new_ids, common_patients)
  
  # Combine
  cor_mat[[cmp_lab]] <- pearson_df %>%
    left_join(kulc_df, by = c("pid_t1", "pid_t2")) %>%
    mutate(
      pid_t1_num = factor(ids_map[as.character(pid_t1)], levels = new_ids),
      pid_t2_num = factor(ids_map[as.character(pid_t2)], levels = new_ids)
    )
}
```

```{r results='asis', fig.width=15, fig.height=6}
# Heatmap: Pearson + Kulczynski (T1/T2)
cmp_heat <- list(c("R_T1", "R_T2"), c("NR_T1", "NR_T2"))
metrics <- c("pearson", "kulczynski")
for (cmp in cmp_heat) {
  
  cmp_lab <- paste(cmp[[1]], "vs", cmp[[2]], sep = "_")
  resp <- ifelse(str_starts(cmp[[1]], "NR"), "NR", "R")

  # Create a tab header (level below the {.tabset} header)
  cat("### ", cmp_lab, "\n\n", sep = "")

  p_heat <- lapply(metrics, function(metric) {
    
    # Create patient labels
    x_labs <- levels(cor_mat[[cmp_lab]]$pid_t2_num)
    y_labs <- levels(cor_mat[[cmp_lab]]$pid_t1_num)
    x_labs[seq_along(x_labs) %% 2 == 0] <- ""
    y_labs[seq_along(y_labs) %% 2 == 0] <- ""
    
    cor_mat[[cmp_lab]] %>%
      ggplot(aes(x = pid_t2_num, y = pid_t1_num, fill = .data[[metric]])) +
      geom_tile() +
      scale_fill_viridis_c(
        option = "plasma",
        # limits = c(0.15, 1),
        breaks = seq(0.2, 0.8, 0.2),
        # oob = scales::squish,
        guide = guide_colorbar(
          barheight = unit(0.85, "npc"), 
          barwidth  = unit(0.5, "cm")
        )
      ) +
      scale_x_discrete(labels = x_labs) +
      scale_y_discrete(labels = y_labs) +
      coord_fixed() +
      theme(
        text = element_text(family = "Montserrat", size = 12),
        panel.grid = element_blank(),
        legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size = 12)
      ) +
      labs(
        title = paste0("Correlation between ", resp, ": T1 vs T2 Epitope Repertoires"),
        subtitle = metric,
        x = paste0(resp, " T2 Ig epitope repertoire (individual count/ID)"),
        y = paste0(resp, " T1 Ig epitope repertoire (individual count/ID)")
      )
  })
  pw <- wrap_plots(p_heat)
  print(pw)
  cat("\n\n")
}
```

## Heatmap - Outlier {.tabset}

I wanted to investigate the outlier in the responder heatmap because it doesn't
make sense to me. First I thought maybe those two samples both have a very low
number of enriched peptides, but that's not the case. Then I thought maybe the
sample labels were swapped, but then the correlation should be symmetric. I now
have the suspicion that patient 23 (T2) wasn't measured and is actually a sample 
from patient PBT313 that was measured twice?

* Patient 23 (T1) doesn't correlate with anything
* For patient PBT313 timepoints correlate with each other
* And they both correlate with patient 23 (T2)

```{r}
checks <- c("T1_PBT313", "T2_PBT313", "T1_23", "T2_23")
hmout_corr <- ps %>%
  mutate(label = paste(timepoint, patientid, sep = "_")) %>%
  filter(label %in% checks) %>%
  select(peptide_id, fold_change, label) %>%
  collect() %>%
  pivot_wider(names_from = "label", values_from = "fold_change") %>% 
  select(all_of(checks))

ggpairs(
  hmout_corr,
  upper = list(continuous = wrap("cor", size = 4, use = "pairwise.complete.obs")),
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.5)),
  diag  = list(continuous = wrap("blankDiag"))
) +
  theme_bw()
```

## Correlation coefficients (Histogram) {.tabset}

```{r}
# Extract correlations of matched/unmatched samples
cor_stats <- list()
for (cmp in names(cor_mat)) {
  matched <- cor_mat[[cmp]] %>%
    filter(pid_t1 == pid_t2) %>%
    pull(pearson)
  
  unmatched <- cor_mat[[cmp]] %>%
    filter(pid_t1 != pid_t2) %>%
    pull(pearson)
  
  cor_stats[[cmp]] <- list(matched = matched, unmatched = unmatched)
}
```

```{r}
y_lim_up <- 180
scale_fct <- 10
col_pal <- c(unmatched = "grey70", matched = group_palette[["R"]])
group <- "Responders"

plot_um_m_corr <- function(cor_um,
                           cor_m,
                           y_lim_up = 200,
                           scale_fct = 10,
                           group = "Group",
                           bins = 40,
                           col_pal = c(um = "grey70", m = "steelblue")) {
  
  df_um <- data.frame(r = cor_um, type = "um")
  df_m <- data.frame(r = cor_m, type = "m")
  
  ggplot() +
    geom_histogram(
      data = df_um,
      aes(x = r, y = after_stat(count), fill = type),
      bins = bins,
      alpha = 0.8
    ) +
    geom_histogram(
      data = df_m,
      aes(x = r, y = after_stat(count) * scale_fct, fill = type),
      bins = bins,
      alpha = 0.8
    ) +
    scale_y_continuous(
      limits = c(0, y_lim_up),
      name = "Unmatched individuals sample correlation",
      sec.axis = sec_axis(
        ~ . / scale_fct, 
        name = "Matched individuals sample correlation"
      )
    ) +
    scale_fill_manual(values = col_pal) +
    labs(
      subtitle = group,
      x = "Pearson correlation of \nlog(fold change) T1 vs T2"
    ) +
    theme_bw(base_size = 11) +
    theme(
      text = element_text(family = "Montserrat"),
      axis.title.y.right = element_text(margin = margin(l = 8), angle = 90, vjust = 0.5),
      axis.title.y.left  = element_text(margin = margin(r = 8)),
      panel.grid = element_blank(),
      legend.position = "none"
    )
}
```

```{r, fig.width=8, fig.height=4}
p1 <- plot_um_m_corr(
  cor_stats$R_T1_vs_R_T2$unmatched,
  cor_stats$R_T1_vs_R_T2$matched,
  y_lim_up = 180,
  group = "Responders",
  col_pal = c(um = phip_palette[[20]], m = group_palette[["R"]])
)

p2 <- plot_um_m_corr(
  cor_stats$NR_T1_vs_NR_T2$unmatched,
  cor_stats$NR_T1_vs_NR_T2$matched,
  y_lim_up = 180,
  group = "Non-responders",
  col_pal = c(um = phip_palette[[20]], m = group_palette[["NR"]])
)

wrap_plots(p1, p2, ncol = 2)
```

## Correlation coefficients (Barplots) {.tabset}

- Do we want statistical significance here? They are all non-significant.
- Add the number of samples that were considered in each comparison?

```{r}
# Extract just the correlations of the matched samples
matched_df <- bind_rows(
  lapply(names(cor_stats), function(nm) {
    vals <- cor_stats[[nm]]$matched
    data.frame(contrast = nm, value = vals)
  })
) %>%
  mutate(
    group = str_extract(contrast, "(R|NR)"),
    time_comp = str_match(contrast, "T(\\d+)_vs_(?:NR|R)_T(\\d+)") |> 
      (\(m) paste0("T", m[,2], "/T", m[,3]))()  
    )

# Compute significance between R/NR at each timepoint comparison
stats_tests <- compare_means(
  value ~ group,
  data  = matched_df,
  group.by = "time_comp",
  method = "wilcox.test",
  p.adjust.method = "BH"
)

# Put everything together
matched_df <- matched_df %>%
  group_by(time_comp, group) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    sd   = sd(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(stats_tests, by = "time_comp") %>%
  mutate(group = factor(group, levels = c("R", "NR")))
```

```{r, fig.width=5, fig.height=5}
ggplot(matched_df, aes(x = time_comp, y = mean, fill = group)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(
    aes(ymin = mean - sd, ymax = mean + sd),
    position = position_dodge(width = 0.7),
    width = 0.2
  ) +
  geom_text(
    aes(x = time_comp, y = 1.1, label = p.signif),
    inherit.aes = FALSE,
    vjust = 0
  ) +
  scale_fill_manual(values = group_palette) +
  labs(
    x = NULL,
    y = "Mean Pearson correlation (matched samples)"
  ) +
  theme_bw(12) +
  theme(
    panel.grid = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```


# Significant Taxa / Annotations {.tabset}

## POP framework {.tabset}

```{r}
pop_df %>%
  filter(category_rank_bh != "not significant") %>%
  arrange(p_raw, p_adj_rank) %>%
  relocate(p_raw, p_adj_rank, percent1, percent2, group1, group2, .after = feature) %>%
  head(10) %>%
  gt()
```

## DELTA framework {.tabset}

```{r}
delta_df %>%
  filter(category_rank_bh != "not significant") %>%
  arrange(p_adj_rank, p_perm) %>%
  relocate(p_adj_rank, p_perm, Z_from_p, .after = feature) %>%
  head(10) %>%
  gt()
```


# Figure 2 {.tabset}

```{r}
make_scatter_flag <- function(pep_tbl, phip_palette) {
  ## Flagellins scatter plot
  # Apply filter
  pep_tbl_prev10 <- pep_tbl %>% filter((prop1 + prop2 >= 0.1))
  # Add peptide library
  peptide_library <- attr(pep_tbl, "prev_meta")$peptide_library
  pep_tbl_peplib <- pep_tbl_prev10 %>%
    left_join(peptide_library, by = c("feature" = "peptide_id"))
  # MW test
  pval_mw <- wilcox.test(ratio ~ is_bac_flagella, data = pep_tbl_peplib)$p.value
  
  # Removing datapoints where is_flagellum is NA, so NAs won't be plotted
  # Alternatively could set NAs to FALSE in peptide_library to plot NAs too
  keep_flag <- peptide_library %>% filter(!is.na(is_bac_flagella))
  pep_tbl_flag <- pep_tbl_prev10 %>% filter(feature %in% keep_flag$peptide_id)
  
  p_static <- scatter_static(
    df = pep_tbl_flag,
    rank = "peptide_id",
    xlab = pep_tbl$group1[1],
    ylab = pep_tbl$group2[1],
    color_by = "is_bac_flagella",
    color_title = paste0("Flagellins (p = ", format_pval(pval_mw), ")"),
    point_size = 2,
    point_alpha = 0,
    font_size = 12
  ) +
    coord_cartesian(xlim = c(-2, 102), ylim = c(-2, 102), expand = TRUE) +
    theme(
      plot.margin = grid::unit(c(12, 12, 12, 12), "pt"),
      text = element_text(family = "Montserrat")
    ) +
    scale_colour_manual(values = c(phip_palette[20], phip_palette[22]))
  
  p_static$data <- p_static$data %>%
    mutate(pt_size = if_else(is_bac_flagella == "yes", 2, 1)) %>%
    arrange(is_bac_flagella)
  
  p_static <- p_static +
    geom_point(
      data = p_static$data,
      aes(x = percent1, y = percent2, colour = is_bac_flagella, size = pt_size),
      alpha = 0.5,
      inherit.aes = FALSE
    ) + 
    scale_size_identity()
  
    return(p_static)
}
```

## Peptide Comparison Between Groups (POP) {.tabset}

- Normally we would test multiple groups not just flagellins, so then the
p-value would have to be corrected for multiple testing.

```{r results='asis', fig.width=5, fig.height=5}
for (nm in names(pep_tbl)) {
  
  # create a tab header (level below the {.tabset} header)
  cat("### ", nm, "\n\n", sep = "")
  
  p <- make_scatter_flag(
    pep_tbl = pep_tbl[[nm]],
    phip_palette = phip_palette
  )
  
  print(p)
  cat("\n\n")
}
```

## Ratios of response by oligo family {.tabset}

- I set an arbitrary prevalence threshold of at least prop1 + prop2 >= 0.1
- For some of the library categories not sure if they are correct?
- Significance calulated with Wilcoxon and BH corrected
- Can't seem to reproduce the same plot. What test?
- How are ratios calculated (to avoid Inf)?

```{r, fig.width=10, fig.height=5}
# Filter by prevalence
pep_tbl_peplib <- pep_tbl$R_T1_vs_NR_T1 %>%
  left_join(peplib, by = c("feature" = "peptide_id")) %>%
  filter((prop1 + prop2 >= 0.1)) %>%
  mutate(
    is_library = TRUE,
    is_bac_flagella = if_else(is_bac_flagella == "yes", TRUE, FALSE)
  )

N1 <- unique(pep_tbl_peplib$N1)
N2 <- unique(pep_tbl_peplib$N2)

# Prepare long format for boxplots by annotation
family_cols <- c(
  "is_library" = "Complete library",
  "is_MPA" = "Microbiome",
  "is_patho" = "Pathogenic strains",
  "is_IgA" = "IgA",
  "is_probio" = "Probiotic",
  "is_bac_flagella" = "Flagellins",
  "is_infect" = "VFDB",
  "is_phage" = "Phages",
  "is_allergens" = "Allergens",
  "is_IEDB_or_cntrl" = "IEDB controls"
)

plot_df <- pep_tbl_peplib %>%
  select(feature, ratio, all_of(names(family_cols))) %>%
  pivot_longer(
    cols = all_of(names(family_cols)),
    names_to = "family",
    values_to = "present"
  ) %>%
  filter(present) %>%
  mutate(family_lab = factor(family_cols[family], levels = family_cols))

# Set colours
oligo_palette <- setNames(phip_palette[1:nlevels(plot_df$family_lab)], family_cols)

# Test each group against reference (Flagellins)
pdat <- compare_means(
  ratio ~ family_lab,
  data = plot_df,
  method = "wilcox.test",
  ref.group = "Flagellins",
  p.adjust.method = "BH"
)

# Remove outliers for plotting (nicer plots)
plot_df <- plot_df %>%
  group_by(family_lab) %>%
  filter(!ratio %in% boxplot.stats(ratio)$out)

# Stacked y positions so brackets don't overlap
max_y <- max(plot_df$ratio, na.rm = TRUE)
pdat <- pdat %>% mutate(y.position = max_y * (1 + 0.1 * row_number()))

# Boxplot (each dot = peptide)
ggplot(plot_df, aes(x = family_lab, y = ratio, fill = family_lab)) +
  geom_boxplot(outlier.shape = NA) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  # scale_fill_manual(values = oligo_palette) +
  ggthemes::scale_fill_tableau("Classic 10") +
  stat_pvalue_manual(
    pdat,
    label = "p.signif",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.01
  ) +
  coord_cartesian(ylim = c(0, 6)) +
  labs(
    title = paste0(
      "Ratios of antibody responses by oligo family T1_PFS12 (", 
      N1, ", ", N2, ")"
    ),
    x = NULL, y = "Ratios of antibody responses in R/NR"
  ) +
  theme_bw(12) +
  theme(
    text = element_text(family = "Montserrat"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    title = element_text(size = 10),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
  )
```

## Ratios of response by class and timepoint - Timepoints {.tabset}

- I can't quite figure out why this plot looks so different here.
- I think there should be a consensus. So far always R/NR and here suddenly PFS12?
- How was significance calculated? Flagellins vs complete library?

```{r, fig.width=9, fig.height=4.5}
flagellins <- pep_tbl_peplib$feature[pep_tbl_peplib$is_bac_flagella == TRUE]

plot_df <- imap_dfr(pep_tbl_xtra, ~ mutate(.x, contrast = .y)) %>%
  filter(feature %in% flagellins) %>%
  separate(
    contrast,
    into = c("var", "value", "t1", "vs", "var2", "value2", "t2"),
    sep = "_"
  ) %>%
  mutate(
    ratio_inv = 1/ratio,
    group_x = factor(
      paste(var, value, sep = "_"), 
      levels = unique(labs$group_lvl)
    ),
    tp_pair = paste(t2, t1, sep = "/"),
    var = factor(var, levels = unique(labs$group_type))
  )
```

```{r, fig.width=10, fig.height=5}
plot_df %>%
  filter(t1 != t2) %>%
  filter(tp_pair %in% c("T2/T1", "T3/T1", "T4/T1")) %>%
  ggplot(aes(x = group_x, y = ratio_inv, fill = tp_pair)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_cartesian(ylim = c(0, 5)) +
  labs(
    title = "Ratios of antibody responses by class and timepoint", 
    x = NULL, 
    y = "Ratio of antibody responses between timepoints", 
    fill = "Timepoint"
  ) +
  scale_fill_manual(values = c(
    "T2/T1" = "#7fc97f",
    "T3/T1" = "#fc8d62",
    "T4/T1" = "#8da0cb"
  )) +
  theme_bw(12) +
  theme(
    text = element_text(family = "Montserrat"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    title = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
  )
```

```{r, fig.width=10, fig.height=5}
plot_df %>%
  filter(t1 != t2) %>%
  filter(var == "PFS12") %>%
  ggplot(aes(x = group_x, y = ratio_inv, fill = tp_pair)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_cartesian(ylim = c(0, 5)) +
  labs(
    title = "Ratios of antibody responses by Class and Timepoint", 
    x = NULL, 
    y = "Ratio of antibody responses between timepoints", 
    fill = "Timepoint"
  ) +
  scale_fill_manual(values = c(
    "T2/T1" = "#7fc97f",
    "T3/T1" = "#fc8d62",
    "T4/T1" = "#8da0cb",
    "T3/T2" = "lightblue",
    "T4/T2" = "maroon",
    "T4/T3" = "#e78ac3"
  )) +
  theme_bw(12) +
  theme(
    text = element_text(family = "Montserrat"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    title = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
  )
```

## Ratios of response by class and timepoint - Class {.tabset}

```{r, fig.width=10, fig.height=5}
plot_df %>%
  filter(t1 == t2) %>%
  ggplot(aes(x = var, y = ratio, fill = t1)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_cartesian(ylim = c(0, 8)) +
  labs(
    title = "Ratios of antibody responses by class and timepoint", 
    x = NULL, 
    y = "Ratio of antibody responses yes/no", 
    fill = "Timepoint"
  ) +
  scale_fill_manual(values = c(
    "T1" = "#7fc97f",
    "T2" = "#fc8d62",
    "T3" = "#8da0cb",
    "T4" = "#e78ac3"
  )) +
  theme_bw(12) +
  theme(
    text = element_text(family = "Montserrat"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    title = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
  )
```


# Additional Plots

- Here the dates are way more extreme than in the original figure. Why?
- Used serum_collection_date-start_io (weeks)

```{r}
meta %>%
  mutate(
    baseline_date = as.Date(start_io, "%d/%m/%Y"),
    timepoint_date = as.Date(serum_collection_date, "%d/%m/%Y"),
    diff_tp_base_week = difftime(timepoint_date, baseline_date, units = "weeks")
  ) %>%
  ggplot(aes(x = timepoint, y = diff_tp_base_week)) +
  geom_boxplot(outliers = TRUE)
```


# Survival analysis {.tabset}

The survival data only has sample IDs (stool samples), no patient IDs. However,
I was able to uniquely match patients based on age, sex and BMI.

```{r}
print("Patients with PhIP-Seq data (T1):")
meta %>% filter(timepoint == "T1") %>% nrow()

print("Patients with PhIP-Seq data (T1) and survival data:")
with_surv <- meta %>%
  mutate(bmi = round(bmi, 3)) %>%
  left_join(
    meta_surv %>% mutate(bmi = round(bmi, 3)), 
    by = c("sex", "age", "bmi")
  ) %>% 
  filter(
    !is.na(os_months),
    timepoint == "T1"
  ) 

with_surv %>% nrow()
```

```{r}
table(with_surv$response)
```
